\speculake: Hiding Malware in Speculative Execution

### 2018-08-01 paper/related-work.tex
working on flow

Previous examples of ``weird machines'' include traditional vulnerabilities such
as buffer overflows, format string exploits and return-to-libc attacks without
function calls~\cite{buffer_overflow, format_string_exploit,
shacham2007geometry}.

Unexpected functionality such as instruction-less computing enables the
demonstration of a Turing complete ``weird machine'' by taking advantage of
memory translation tables, page-faults, and hardware exception handling to
effectively compute arithmetic and logical operations without the processor
executing any instructions~\cite{bangert2013page}.

Features in modern processors such as multiple threads sharing system resources
and optimizations done across the process isolation boundary leads to
opportunities for ``weird machines''~\cite{d2015exploiting}.

@@

\subsection{Red-Pills}
Hardware minutia have also been used to identify and fingerprint execution
environments---in malware this is known as a red-pill~\cite{red-pill} and is
typically used by malware to detect when it is under emulation or a debugging
environment~\cite{lindorfer2011detecting, balzarotti2010efficient,
paleari2009fistful} attempting to circumvent malware detectors.

To address red-pills, systems often employ measures to detect them.
``Bare-metal'' emulation of malware~\cite{kirat2011barebox}, methods to
determine if malware is performing red-pill checks to avoid emulation
environments~\cite{kirat2014barecloud}, and symbolic execution can be used to
combat red-pills by finding environmental triggers in malicious
binaries~\cite{schwartz2010all}.

\speculake undertakes to avoid detection, and is a new class of red-pill. Unlike
previous red-pills, \speculake complicates analysis and impedes red-pill
detectors with the use a trigger program and speculative execution. The use of a
separate and seemingly benign trigger program influenced by the adversary
enables the \speculake malware to limit detection by acting benign until
triggered to influence the branch predictor. Additionally, as an \speculake
attack is carried out via speculative execution, the instructions executed and
speculative state are not committed to by the processor and are thus obscured
from analysts.

Similarly, break-points (and other debugging techniques) will not aid in
unravelling \speculake malware as break-points interrupt the control flow of a
program and prevent speculative execution. Symbolic execution will also fail to
find the speculative gadgets as they are never on the (committed) execution
pathway. This allows \speculake to effectively act as a red-pill while avoiding
detection as such.

@@

The branch predictor has also been used as a covert
channel~\cite{evtyushkin2016understanding}. Previous work has examined the
capability of two colluding programs to share sensitive information between each
other using the branch predictor by timing how long a spy process executes taken
(not-taken) branches~\cite{evtyushkin2015covert}.

\speculake exemplifies a new covert channel using the branch predictor in which
only one side of the channel is nefarious and the other is innocent. \speculake
takes advantage of a branch predictor side-channel in \texttt{OpenSSL} and uses
it as a covert channel. \texttt{OpenSSL} is not a malicious colluding program in
this case but rather a program whose execution pathway can be influenced by
attackers based on selected cipher suites. This allows malware to use the branch
predictor as an attacker controlled covert channel between an unsuspecting
trigger program and \speculake malware.

>>>>

Recently there has been a trend of features in modern processors such as
multiple threads sharing system resources and optimizations done across the
process isolation boundary which lead to opportunities for ``weird
machines''~\cite{d2015exploiting}. In particular, previous examples of ``weird
machines'' include traditional vulnerabilities such as buffer overflows, format
string exploits and return-to-libc attacks without function
calls~\cite{buffer_overflow, format_string_exploit, shacham2007geometry}.
Additionally, unexpected functionality such as instruction-less computing
enables the demonstration of a Turing complete ``weird machine'' by taking
advantage of memory translation tables, page-faults, and hardware exception
handling to effectively compute arithmetic and logical operations without the
processor executing any instructions~\cite{bangert2013page}.

@@

\subsection{Red-Pills}
Hardware minutia have also been used to identify and fingerprint execution
environments---in malware this is known as a red-pill~\cite{red-pill} and is
typically used by malware to detect when it is under emulation or a debugging
environment~\cite{lindorfer2011detecting, balzarotti2010efficient,
paleari2009fistful} attempting to circumvent malware detectors. To address
red-pills, systems often employ measures to detect them. ``Bare-metal''
emulation of malware~\cite{kirat2011barebox}, methods to determine if malware is
performing red-pill checks to avoid emulation
environments~\cite{kirat2014barecloud}, and symbolic execution can be used to
combat red-pills by finding environmental triggers in malicious
binaries~\cite{schwartz2010all}.

Analogously, \speculake undertakes to avoid detection. Unlike previous
red-pills, \speculake complicates analysis and impedes malware detectors with
the use a trigger program and speculative execution. The use of a separate and
seemingly benign trigger program influenced by the adversary enables the
\speculake malware to limit detection by acting benign until triggered to
influence the branch predictor (while still appearing innocuous).

Additionally, as an \speculake attack is carried out via speculative execution,
the instructions executed and speculative state are not committed to by the
processor and are thus obscured from analysts. Similarly, symbolic execution
will fail to find the speculative gadgets as they are never on the (committed)
execution pathway. More direct analysis such as with break-points (and other
debugging techniques) will not aid in unravelling \speculake malware as
break-points interrupt the control flow of a program and prevent speculative
execution. This allows \speculake to effectively act as a red-pill while
avoiding detection as such.

@@

\speculake uses the branch predictor as a covert channel. The branch predictor
has been considered as a covert channel
before~\cite{evtyushkin2016understanding}. For example, the branch predictor can
be used by two colluding programs to share sensitive information between each
other by timing how long a spy process executes taken (not-taken)
branches~\cite{evtyushkin2015covert}.

While the branch predictor has been used as a covert channel between two
colluding malicious programs, \speculake exemplifies a new covert channel using
the branch predictor in which only one side of the channel is nefarious and the
other is innocent. \speculake takes advantage of a branch predictor side-channel
in \texttt{OpenSSL} and uses it as a covert channel. \texttt{OpenSSL} is not a
malicious colluding program in this case but rather a program whose execution
pathway can be influenced by attackers based on selected cipher suites. This
allows malware to use the branch predictor as an attacker controlled covert
channel between an unsuspecting trigger program and \speculake malware.
